---
title: "Stats Project"
author: "Faith Gibson"
output: html_notebook
---

# Stats Project

## Objectives for this project

To complete this project you’ll need to do a few things within this file. As you go through the notebook, you will have further instruction on how to complete these objectives.

1. Go through this notebook, reading along.

2. Fill in empty or incomplete code chunks when prompted to do so.

3. Run each code chunk as you come across it by clicking the tiny green triangles at the right of each chunk. You should see the code chunks print out various output when you do this.

4. At the very top of this file, put your own name in the `author:` place. Currently it says `"DataTrail Team"`. Be sure to put your name in quotes.

5. In the `Conclusions` section, write up responses to each of these questions posed here.

6. When you are satisfied with what you've written and added to this document you'll need to save it. In the menu, go to `File` > `Save`. Now the `nb.html` output resulting file will have your new output saved to it.

7. Open up the resulting `countries_stats.nb.html` file and click `View in Web Browser`. Does it look good to you? Did all the changes appear here as you expected.

8. Upload your `Rmd` and your `nb.html` to your assignment folder (this is something that will be dependent on what your instructors have told you -- or if you are taking this on your own, just collect these projects in one spot, preferably a Google Drive)!

9. Pat yourself on the back for finishing this project!

## The goal of this analysis

<Write here what the goal of this analysis is. What question are we trying to answer?>

## Set up

Let's load these packages for use.

```{r}
## you can add more, or change...these are suggestions
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(corrplot)
library(knitr)
```

## Set up directories

Here we are going to make a data directory if it doesn't already exist.

```{r}
if (!dir.exists("data")) {
  dir.create("data")
}
if (!dir.exists("plots")) {
  dir.create("plots")
}
```


## Get the data

In a previous project in `03_Cleaning_Data`, we cleaned a dataset about countries and saved this to an RDS file.
Read in this file from its location in the `03_Cleaning_Data` folder. **Hint** you will need to use a `..` to backout of this current folder and reference the parent folder of `05_Get_Stats`.

```{r, solution = TRUE}
readRDS("../03_Cleaning_Data/data/countries_df.rds")
```

Use functions we have discussed previously to see remind yourself of what variables are in `countries_df`.

```{r, solution = TRUE}
colnames(countries_df)
```

### Cleaning the datasets

We're going to select down to columns we are interested in .

```{r}
countries_df <- countries_df %>%
  select(country,
         pop = population_in_thousands_2017,
         pop_density = population_density_per_km2_2017,
         sex_ratio = sex_ratio_m_per_100_f_2017,
         gdp = gdp_per_capita_current_us,
         labour_force_participation_female_male_pop_percent,
         urban_perc = urban_population_percent_of_total_population,
         growth_rate = population_growth_rate_average_annual_percent,
         fertility = fertility_rate_total_live_births_per_woman,
         life_exp_female = life_expectancy_at_birth_females,
         life_exp_male =life_expectancy_at_birth_males,
         infant_mortality = infant_mortality_rate_per_1000_live_births,
         seats_held_women = seats_held_by_women_in_national_parliaments_percent,
         pop_using_improved_drinking_water_urban_rural_percent)
```


There are some variables here that need to be split up into two columns much like we did before in `03_Cleaning_the_Data`.

Repeat the steps we used in that project for the following columns:

`labour_force_participation_female_male_pop_percent`
`pop_using_improved_drinking_water_urban_rural_percent`     


```{r, solution = TRUE}
countries_df <- tidyr::separate(countries_df, labour_force_participation_female_male_pop_percent, into = c("labour_force_participation_females", "labour_force_participation_males"), sep = "/")
```

```{r, solution = TRUE}
countries_df <- tidyr::separate(countries_df, pop_using_improved_drinking_water_urban_rural_percent, into = c("pop_using_improved_drinking_water_urban", "pop_using_improved_drinking_water_rural"), sep = "/")
```


Now all of the data in our set, except for `country`, should be numeric, but it isn't necessarily. Use `mutate_at` but exclude `country` and try to use as.numeric on all the other variables. **Hint** you will need an argument that specifies `vars(-country)` to exclude the `country` column.

```{r, solution = TRUE}
countries_df <- countries_df %>%
  mutate_at(vars(-country), as.numeric)
```

Save this version of your data to an RDS file in this project.

```{r, solution = TRUE}
saveRDS(countries_df, "data/countries_df.rds")
```

## Plot the data

Let's explore what kinds of distributions these variables have?

Use ggplot2::geom_density() to do this for at least 5 variables of your choice in `countries_df`.

```{r, solution = TRUE}
countries_df %>% ggplot(aes(pop_density)) + geom_density()
```

```{r, solution = TRUE}
countries_df %>% ggplot(aes(urban_perc)) + geom_density()
```

```{r, solution = TRUE}
countries_df %>% ggplot(aes(labour_force_participation_females)) + geom_density()
```

```{r, solution = TRUE}
countries_df %>% ggplot(aes(life_exp_male)) + geom_density()
```

```{r, solution = TRUE}
countries_df %>% ggplot(aes(infant_mortality)) + geom_density()
```

```{r, solution = TRUE}
countries_df %>% ggplot(aes(pop_using_improved_drinking_water_rural)) + geom_density()
```

## Run stats  

For this project, we are going to run some exploratory stats.

Use the `cor()` function to create a correlation matrix of all the numeric variables in `countries_df`

```{r, solution = TRUE}
countries_cor <- countries_df %>% 
  rename(labour_force_females = labour_force_participation_females,
         labour_force_males = labour_force_participation_males,
         improved_water_urban = pop_using_improved_drinking_water_urban,
         improved_water_rural = pop_using_improved_drinking_water_rural)

countries_cor <- cor(countries_cor[sapply(countries_cor, is.numeric)])
```

Make a table out of this correlation matrix.

```{r, solution = TRUE}
kable(countries_cor)
```

Use the corrplot::corrplot() function to make a correlation matrix from these data.

```{r, solution = TRUE}
corrplot::corrplot(countries_cor,
                   col = COL2("PRGn", 200),
                   tl.col = "black",
                   na.label = "square",
                   na.label.col = "lightgrey")
```

What do you think about the interaction and groupings of these variables? Is there anything surprising?

```{r}
# Assuming I did everything correctly, I am surprised that there isn't much correlation between the majority of variables in this data set. It makes sense that there would be, at least, a small correlation between population-related variables and urban percentage, but I would've expected to see, let's say, a positive correlation between females in the labour force and seats held by women. Or, a negative correlation between infant mortality rates and percentage of urban/rural areas using improved drinking water.
```

Save your plot using the following code.

```{r, solution = TRUE}
# where is the code :fear:
# stackoverflow to the rescue??

countries_plot <- {
                    corrplot::corrplot(countries_cor,
                                       col = COL2("PRGn", 200),
                                       tl.col = "black",
                                       na.label = "square",
                                       na.label.col = "lightgrey");
                    recordPlot()
                  }

ggsave("plots/countries_plot.png", plot = replayPlot(countries_plot))

```

### Further exploration

Based on the plot above, do some statistics and a plot to further explore the relationship of two of the variables in `countries_df`.

```{r}
countries_df %>% ggplot(aes(pop_density, urban_perc)) +
                 geom_point() +
                 labs(title = "Effect of Population Density on Urbanization",
                      x = "Population Density",
                      y = "Urban Area Percentage") +
                 theme_excel()
```

```{r}
wilcox.test(countries_df$pop_density, countries_df$urban_perc)
```

### Conclusion

Write up your thoughts about this data science project here and answer the following questions:

**- What have you learned about the relationship of the two variables you chose?**

Even though the p-value would be considered statistically significant (1.375e-05 = 0.00001375), there are several outliers that could be influencing the result to be that small, looking at the scatter plot, there doesn't appear to be a strong relationship between urbanization (urban_perc) and population density, which reflects what is shown in the correlation plot.

**- Was anything surprising or confusing?**

It was very confusing to see that there weren't any strong correlations between all the different variables, not counting when a variable was being compared with itself. Maybe I did something wrong when making the correlation matrix and/or using kable?

**- What alternative hypothetical explanations are there for the observed relationships you are seeing? (Remember about third variable problem means you can't say things are causal by correlations)**

Honestly cannot answer this, given the above.

**- In a hypothetical future study, what questions would you want to explore and what kind of data would you want to use to do this exploration?**

I would want to use the variables pertaining to life expectancy rates, infant mortality rate, and improved drinking water percentages to explore how having access to cleaner ("improved") drinking water affects the estimated life span of humans.

## Print out session info

Session info is a good thing to print out at the end of your notebooks so that you (and other folks) referencing your notebooks know what software versions and libraries you used to run the notebook.

```{r}
sessionInfo()
```
